---
title: "PR2 - Use of the R pr2database data package"
author: "Daniel Vaulot"
date: '`r format(Sys.time(), "%d %m %Y")`'
output:
  md_document:
    variant: markdown_github
---

```{r readme-setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "img/"   # Directories where image are saved
)
is_online = curl::has_internet()
```

The PR2 database is now provided as a R package

# Installation

Install from the GitHub web site using the devtools package

```{r, eval=FALSE, message=FALSE}
install.packages(devtools)
devtools::install_github("vaulot/pr2database")
```
```
* installing *source* package 'pr2database' ...
** R
** data
*** moving datasets to lazyload DB
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
  converting help for package 'pr2database'
    finding HTML links ... fini
    pr2                                     html  
** building package indices
** testing if installed package can be loaded
*** arch - i386
*** arch - x64
* DONE (pr2database)
In R CMD INSTALL
```


# Loading the database

The PR2 database is provided as a data frame (or a tibble). This is a join between the following tables:
* pr2_main
* pr2_taxonomy
* pr2_sequence
* pr2_metadata

```{r}
library("pr2database")

data("pr2")

# List of the different columns available - see the help of the package for information on each field

colnames(pr2)

```

# Working with the database


## Install and load the libraries
The following examples makes use of the specifc R libraries

Install the libraries

```{r, eval=FALSE}
install.packages("dplyr")      # For filtering the data
install.package("ggplot2")     # To plot data
install.package("maps")        # To plot maps

source("https://bioconductor.org/biocLite.R")  # This package is on Bioconductor
biocLite("Biostrings")         # To save fasta files
```

Load the libraries

```{r}
  library(dplyr)
  library(ggplot2)    # For plots
  library(Biostrings) # To save fasta files
```


## Selecting sequences from a specific taxon

Let us select all the available sequences for the Mamiellophyceae _Ostreococcus_

```{r}

  # Filter only the sequences for which the column genus contains Ostreococcus
  pr2_ostreo <- pr2 %>% dplyr::filter(genus == "Ostreococcus")

  # Select only the columns of interest
  pr2_ostreo <- pr2_ostreo %>% dplyr::select( genbank_accession, species, 
                                              pr2_sample_type, gb_strain, gb_clone, 
                                              pr2_latitude, pr2_longitude, 
                                              sequence_length, sequence  )
  
  pr2_ostreo
```

## Exporting the sequences to fasta

We will save the _Ostreococcus_ sequences to a FASTA file. This is easy done with the bioconductor package BioStrings.

```{r}

  # Importing the sequence in a Biostring set 

  seq_ostreo <- Biostrings::DNAStringSet(pr2_ostreo$sequence)

  # Constructing the name of each sequecne (the first line of the fasta file)
  # using the genbank accession, species name, strain name and clone name

  names(seq_ostreo) <- paste(pr2_ostreo$genbank_accession, pr2_ostreo$species,
                             "strain",pr2_ostreo$gb_strain,
                             "clone",pr2_ostreo$gb_clone, 
                              sep="|")

  # Displaying the Biostring set
  seq_ostreo
    
  # Saving the sequences as a fasta file
  Biostrings::writeXStringSet(seq_ostreo, "examples/pr2_ostreo.fasta", width = 80)

```

The fasta file will look as follows

```
>AF525872|Ostreococcus_lucimarinus|strain|NA|clone|UEPACIp5
ACCTGGTTGATCCTGCCAGTAGTCATATGCTTGTCTCAAAGATTAAGCCATGCATGTCTAAGTATAAGCGTTATACTGTG
AAACTGCGAATGGCTCATTAAATCAGCAATAGTTTCTTTGGTGGTGTTTACTACTCGGATAACCGTAGTAATTCTAGAGC
TAATACGTGCGTAAATCCCGACTTCGGAAGGGACGTATTTATTAGATAAAGACCG...
>EU562149|Ostreococcus_lucimarinus|strain|NA|clone|IND2.6
AAAGCTCGTAGTCGGATTTTGGCTGAGAACGGTCGGTCCGCCGTTAGGTGTGCACTGACTGGTCTCAGCTTCCTGGTGAG
GAGGTGTGCTTCATCGCCACTTAGTCACCGTGGTTACTTTGAAAAAATTAGAGTGTTCAAAGCGGGCTTACGCTTGAATA
TATTAGCATGGAATAACACCATAGGACTCCTGTCCTATTTCGTTGGTCTCGGGACGGGAGTAATGATTAAGATGAACAGT
TGGGGGCATTCGTATTTCATTGTCAGAGGTGAAATTCTTGGATTT...
>AY425309|Ostreococcus_lucimarinus|strain|NA|clone|RA010412.39
GCCAGTAGTCATATGCTTGTCTCAAAGATTAAGCCATGCATGTCTAAGTATAAGCGTTATACTGTGAAACTGCGAATGGC
TCATTAAATCAGCAATAGTTTCTTTGGTGGTGTTTACTACTCGGATAACCGT...
```

## Doing an histogram of the sequence length

```{r sequence_histogram}
  ggplot(pr2_ostreo) + 
    geom_histogram(aes(sequence_length), binwidth = 50, fill="blue") + 
    xlim(0,2000) + xlab("Sequence length") + ylab("Number of sequences") + 
    ggtitle("Ostreococcus lucimarinus")
```


## Drawing a map of sequence locations

```{r sequence_map}
  library(maps)
  world <- map_data("world")

  ggplot() + 
    geom_polygon(data = world, aes(x=long, y = lat, group = group), fill="grey") + 
    coord_fixed(1.3) +
    geom_point(data=pr2_ostreo, aes(x=pr2_longitude, y=pr2_latitude), fill="blue", size=2, shape=21) + 
    ggtitle("Ostreococcus lucimarinus")
```

