---
title: "PR2"
author: Daniel Vaulot
date: March 22, 2005
output:
  md_document:
    variant: markdown_github
---

```{r echo=FALSE, fig.align="center", fig.cap="", out.width='50%'}
knitr::include_graphics('/img/pr2_logo')
```

# Protist Ribosomal Reference database (PR2)

[![Release](https://img.shields.io/badge/release-4.10.0-blue.svg) ![Date](https://img.shields.io/badge/date-7%20March%202018-lightgrey.svg)](https://github.com/vaulot/pr2_database/releases)  [
![Twitter URL](https://img.shields.io/twitter/url/http/shields.io.svg?style=social)
](https://twitter.com/intent/tweet?text=PR2%20database&url=https://github.com/vaulot/pr2_database&hashtags=PR2database)

## SSU rRNA gene database

The Protist Ribosomal Reference database (PR2) provides a unique access to eukaryotic small sub-unit (SSU) ribosomal RNA and DNA sequences, with curated taxonomy. The database mainly consists of nuclear-encoded protistan sequences. However, metazoans, land plants, macrosporic fungi are also included because they are useful for the analysis of high-troughput sequencing data sets. Introns and putative chimeric sequences have been also carefully checked. Taxonomic assignation of sequences consists of eight unique taxonomic fields.

Since the original web site (http://ssu-rrna.org/pr2) does not exist any more, we are proposing updated version of PR2  and provide the data in a variety of file formats to use for annotating metabarcodes.


## Current version
* Current version : 4.10.0
* Last update : 7 March 2018
* DOI : https://doi.org/10.6084/m9.figshare.5913181
* [Link to latest release](https://github.com/vaulot/pr2_database/releases)
* Manual : https://github.com/vaulot/pr2_database/wiki

## Basic statistics
* [PR2 statistics](https://vaulot.github.io/pr2/PR2_analysis.html)

## Report issues
* Please report any issue on [GitHub](https://github.com/vaulot/pr2_database/issues)

## Contact
Daniel VAULOT, Laure GUILLOU and Fabrice NOT
DIPO team, Plankton Group, UMR 7144 CNRS-UPMC
Station Biologique,
Place G. Tessier
29680 Roscoff FRANCE
email: vaulot@sb-roscoff.fr / vaulot@gmail.com

## Contributors
- Tristan Biard (Wimereux Station)
- Margot Tragin (Roscoff)
- Bente Edvardsen (U of Oslo)
- Charles Bachy and Wei Ting Chen (Roscoff)
- Solenn Mordret (Naples)
- Adriana Lopes dos Santos (Nanyang University, Singapore)
- Vittorio Boscaro (UBC, Vancouver)

## Please cite
* Guillou, L., Bachar, D., Audic, S., Bass, D., Berney, C., Bittner, L., Boutte, C. et al. 2013. The Protist Ribosomal Reference database (PR2): a catalog of unicellular eukaryote Small Sub-Unit rRNA sequences with curated taxonomy. Nucleic Acids Res. 41:D597–604.

# Related Databases

## Eukref
```{r echo=FALSE, fig.align="center", fig.cap="", out.width='30%'}
knitr::include_graphics('/img/eukref_logo')
```  
PR2 will be incorporating annotations made by the [EukRef](http://eukref.org/) project as deposited in their [GitHub repository](https://github.com/eukref/curation).

## PhytoRef (16S plastid)
```{r echo=FALSE, fig.align="center", fig.cap="", out.width='30%'}
knitr::include_graphics('/img/eukref_logo')
``` 
Plastid 16S rRNA sequences are yet not included into PR2.  These sequences are available from the the PhytoRef database available from [Figshare](https://figshare.com/articles/PhytoREF_a_reference_database_of_the_plastidial_16S_rRNA) or [PhytoRef](http://phytoref.sb-roscoff.fr/).
