---
title: "PR2 statistics"
author: "Daniel Vaulot"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{tutorial: PR2 statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  prompt=FALSE,
  tidy=FALSE,
  comment = "#>" 
#  fig.path = "img/"   # Directories where image are saved
)

# Notes
# - cannot use rmarkdown::html_vignette for the moment (maybe with R version 4.0)
# - use build_vignettes() to build the vignettes
# - need to build source package to have the vignettes
```

---
title: "PR2 statistics"
author: vaulot
date: '2018-08-10'
lastmod: "2019-11-21"
categories: ["documentation"]
tags: ["Database", "Statistics", "R-package"]
summary: "Quick facts about PR2."
---


```{r message=FALSE, warning=FALSE}
  library("ggplot2")
  library("dplyr")
```


```{r message=FALSE, warning=FALSE}


  library("pr2database")
  data(pr2)
  packageVersion("pr2database")

  pr2_photo <- pr2 %>% filter((division %in% c("Chlorophyta", "Dinophyta", "Cryptophyta",
                                                 "Haptophyta", "Ochrophyta", "Cercozoa")) &
                                              !(class %in% c("Syndiniales", "Sarcomonadea")))
  pr2_ref <- pr2 %>% filter(!is.na(reference_sequence))
```


# PR2 fields

```{r}
colnames(pr2)
```


# Basic statistics

## All taxa

Total number of PR2 sequences : `r nrow(pr2)`

```{r}
#  stringr::str_c("Number of sequences : ", nrow(pr2))

  pr2_taxa <- pr2 %>% select(kingdom:genus, species) %>% summarise_all(funs(n_distinct(.)))
  knitr::kable(pr2_taxa, caption="Number of taxa - all sequences")
```

## Photosynthetic protists

Number of photosynthetic protist sequences : `r nrow(pr2_photo)`

```{r}
#  stringr::str_c("Number of sequences (active only) : ", nrow(pr2_photo))

  pr2_taxa <- pr2_photo %>% select(kingdom:genus, species) %>% summarise_all(funs(n_distinct(.)))
  knitr::kable(pr2_taxa, caption="Number of taxa - photosynthetic protist sequences")
```

## Reference sequences

_Reference sequences are a subset of PR2 representative of taxonomic groups._

Number of reference sequences : `r nrow(pr2_ref)`

```{r}
#  stringr::str_c("Number of reference sequences : ", nrow(pr2_ref))

  pr2_taxa <- pr2_ref %>% select(kingdom:genus, species) %>% summarise_all(funs(n_distinct(.)))
  knitr::kable(pr2_taxa, caption="Number of taxa - Reference sequences")
```

# Sequence length

```{r sequence length}
  ggplot(pr2) + geom_histogram(aes(sequence_length), binwidth = 100, fill="blue") +
                xlim(0,3000) + xlab("PR2 sequence length") + ylab("Number of sequences") + ggtitle("All sequences")
  ggplot(pr2_ref) + geom_histogram(aes(sequence_length), binwidth = 100, fill="blue") +
                xlim(0,3000) + xlab("PR2 sequence length") + ylab("Number of sequences") + ggtitle("Reference sequences")
```

# Taxonomic composition

```{r}
  pr2.env <- new.env()

  pr2.env$taxo_levels = c("kingdom", "supergroup", "division", "class", "order", "family", "genus", "species")
  pr2.env$taxo_levels_number =length(pr2.env$taxo_levels)

pr2_treemap <- function(pr2, taxo_rank) {

  # Define the levels
  level1 = pr2.env$taxo_levels[taxo_rank]
  level2 = pr2.env$taxo_levels[taxo_rank+1]
  # Group
  pr2_class <- pr2 %>%
                group_by_(level1,level2) %>%
                summarise(sequence_number= n())

  # Do a simple treemap
  treemap::treemap(pr2_class, index=c(level1,level2),
        vSize="sequence_number",
        title="",asp=1, lowerbound.cex.labels= 0.2, fontsize.labels = 12,
        palette="Blues",
        format.legend = list(scientific = FALSE, big.mark = " "))
}

```


## Division level

### All groups

```{r treemaps}
  pr2_treemap(pr2,3)
```

### Reference sequences

```{r}
  pr2_treemap(pr2_ref,3)
```

### Photosynthetic protists

```{r}
  pr2_treemap(pr2_photo,3)
```


## Genera most represented

### All taxa

```{r genera histo}
  pr2_genus <- pr2 %>% group_by(class, genus) %>% count() %>% ungroup() %>% top_n(30)

  ggplot(pr2_genus) +
    geom_col(aes(x=forcats::fct_reorder(stringr::str_c(class,"-",genus), n), y=n)) +
    coord_flip() +
    ggtitle("Most represented genera - all") +
    xlab("Genera") + ylab("Number of sequences")

```

### Reference sequences

```{r}
  pr2_genus <- pr2_ref %>% group_by(class, genus) %>% count() %>% ungroup() %>% top_n(30)

  ggplot(pr2_genus) +
    geom_col(aes(x=forcats::fct_reorder(stringr::str_c(class,"-",genus), n), y=n)) +
    coord_flip() +
    ggtitle("Reference sequences") +
    xlab("Genera") + ylab("Number of sequences")

```

### Only photosynthetic protists

```{r}

  pr2_genus <- pr2_photo %>% group_by(class, genus) %>% count() %>% ungroup() %>% top_n(30)

  ggplot(pr2_genus) +
    geom_col(aes(x=forcats::fct_reorder(stringr::str_c(class,"-",genus), n), y=n)) +
    coord_flip() +
    ggtitle("Most represented genera - only photosynthetic protists") +
    xlab("Genera") + ylab("Number of sequences")
```


# World sequence distribution

```{r}

map_get_world <- function(resolution="coarse"){
  worldMap <- rworldmap::getMap(resolution = resolution) # Change to "coarse" for global maps / "low" for regional maps
  world.points <- fortify(worldMap)
  world.points$region <- world.points$id
  world.df <- world.points[,c("long","lat","group", "region")]
  }


map_world <- function(color_continents = "grey80", color_borders = "white", resolution = "coarse") {

  # Background map using the maps package
  # world.df <- map_data("world")

  world.df <- map_get_world(resolution)

  map <- ggplot() +
    geom_polygon(data = world.df, aes(x=long, y = lat, group = group), fill=color_continents, color=color_borders) +
    # scale_fill_manual(values= color_continents , guide = FALSE) +
    scale_x_continuous(breaks = (-4:4) * 45) +
    scale_y_continuous(breaks = (-2:2) * 30) +
    xlab("Longitude") + ylab("Latitude") +
    coord_fixed(1.3) +
    theme_bw()
    # species_map <- species_map + coord_map ()  # Mercator projection
    # species_map <- species_map + coord_map("gilbert") # Nice for the poles
  return(map)
  }
```


## All taxa

```{r map}
 map_world() + geom_point(data=pr2, aes(x=pr2_longitude, y=pr2_latitude), fill="blue", size=2, shape=21) +
               ggtitle("PR2 - all sequences")  
```

## Photosynthetic protists

```{r}
 map_world() + geom_point(data=pr2_photo, aes(x=pr2_longitude, y=pr2_latitude), fill="red", size=2, shape=21) +
               ggtitle("PR2 - photosynthetic protists sequences")  
```

# Number of sequences per country and ocean

Version 4.12.0 incorporate better geo-localisation using the approach pioneered by [Margaret Mars Brisbin](https://twitter.com/MargaretBrisbin)

## Number of sequences per country of origin

```{r sequence_country}

  countries <- pr2 %>%
    count(pr2_country) %>%
    arrange(-n) %>%
    filter(!is.na(pr2_country) & n > 500)

  ggplot(countries, aes(x = reorder(pr2_country, n), y = n)) +
    geom_col() +
    coord_flip() +
    xlab("")  + ylab("Number of PR2 sequences")
```

## Number of sequences per ocean of origin

```{r sequence_oceans}

  oceans <- pr2 %>%
    count(pr2_ocean) %>%
    arrange(-n) %>%
    filter(!is.na(pr2_ocean))

  ggplot(oceans, aes(x = reorder(pr2_ocean, n), y = n)) +
    geom_col() +
    coord_flip() +
    xlab("")  + ylab("Number of PR2 sequences")
```
