---
title: "PR2 statistics"
author: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial: PR2 statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  prompt=FALSE,
  tidy=FALSE,
  comment = "#>" 
#  fig.path = "img/"   # Directories where image are saved
)

# Notes
# - cannot use rmarkdown::html_vignette for the moment (maybe with R version 4.0)
# - use devtools::build_vignettes() to build the vignettes
# - need to build source package to have the vignettes
```


```{r message=FALSE, warning=FALSE}
  library("ggplot2")
  library("dplyr")
  library("knitr")
  library("forcats")
  library("stringr")
  library("rworldmap")
  library("treemapify")
```


```{r message=FALSE, warning=FALSE}


  library("pr2database")
  data(pr2)
  packageVersion("pr2database")

  pr2_photo <- pr2 %>% filter((division %in% c("Chlorophyta", "Dinophyta", 
                                               "Cryptophyta",
                                               "Haptophyta", "Ochrophyta")) &
                              !(class %in% c("Syndiniales", "Sarcomonadea")))
  pr2_ref <- pr2 %>% filter(!is.na(reference_sequence))
```


## PR2 fields

```{r}
colnames(pr2)
```


## Basic statistics

### All taxa

Total number of PR2 sequences : **`r nrow(pr2)`**

```{r}

pr2_taxa <- pr2 %>% select(kingdom:genus, species) %>% 
  summarise_all(funs(n_distinct(.)))
knitr::kable(pr2_taxa, caption="Number of taxa - all sequences")
```

### Photosynthetic protists

Number of photosynthetic protist sequences : **`r nrow(pr2_photo)`**

```{r}
pr2_taxa <- pr2_photo %>% select(kingdom:genus, species) %>% 
  summarise_all(funs(n_distinct(.)))
knitr::kable(pr2_taxa, caption="Number of taxa - photosynthetic protist sequences")
```

### Reference sequences

_Reference sequences are a subset of PR2 representative of taxonomic groups._

Number of reference sequences : **`r nrow(pr2_ref)`**

```{r}

pr2_taxa <- pr2_ref %>% select(kingdom:genus, species) %>% 
  summarise_all(funs(n_distinct(.)))
knitr::kable(pr2_taxa, caption="Number of taxa - Reference sequences")
```

## Sequence length

```{r sequence length}
ggplot(pr2) + geom_histogram(aes(sequence_length), binwidth = 100, fill="blue") +
  xlim(0,3000) + xlab("PR2 sequence length") + 
  ylab("Number of sequences") + 
  ggtitle("All sequences")
ggplot(pr2_ref) + geom_histogram(aes(sequence_length), binwidth = 100, fill="blue") +
  xlim(0,3000) + 
  xlab("PR2 sequence length") + 
  ylab("Number of sequences") + 
  ggtitle("Reference sequences")
```

## Taxonomic composition

```{r}

pr2_treemap <- function(pr2, level1, level2) {


  # Group
  pr2_class <- pr2 %>%
    count({{level1}},{{level2}}) %>% 
    filter(!is.na(division)) %>%
    ungroup()

  # Do a treemap
  
  ggplot(pr2_class, aes(area = n, 
                        fill = {{level2}}, 
                        subgroup = {{level1}}, 
                        label = {{level2}})) +
           treemapify::geom_treemap()
  
  ggplot(pr2_class, aes(area = n, 
                        fill= {{level1}}, 
                        subgroup = {{level1}}, 
                        label = {{level2}})) +
    treemapify::geom_treemap() +
    treemapify::geom_treemap_text(colour = "white", place = "centre", grow = TRUE) +
    treemapify::geom_treemap_subgroup_border() +
    treemapify::geom_treemap_subgroup_text(place = "centre", grow = T, 
                                           alpha = 0.5, colour = "black", 
                                           min.size = 0) +
    theme_bw() +
    scale_color_brewer() +
    guides(fill = FALSE)
           
}

```


### Division level

#### All groups

```{r}
pr2_treemap(pr2, division, class)
```

#### Reference sequences

```{r}
pr2_treemap(pr2_ref, division, class)
```

#### Photosynthetic protists

```{r}
pr2_treemap(pr2_photo,division, class)
```


### Genera most represented

#### All taxa

```{r genera histo}
pr2_genus <- pr2 %>% group_by(class, genus) %>% 
  count() %>% 
  ungroup() %>% 
  top_n(30)

ggplot(pr2_genus) +
  geom_col(aes(x=forcats::fct_reorder(stringr::str_c(class,"-",genus), n), y=n)) +
  coord_flip() +
  ggtitle("Most represented genera - all") +
  xlab("Genera") + ylab("Number of sequences")

```

#### Reference sequences

```{r}
pr2_genus <- pr2_ref %>% 
  group_by(class, genus) %>% 
  count() %>% ungroup() %>% 
  top_n(30)

ggplot(pr2_genus) +
  geom_col(aes(x=forcats::fct_reorder(stringr::str_c(class,"-",genus), n), y=n)) +
  coord_flip() +
  ggtitle("Reference sequences") +
  xlab("Genera") + 
  ylab("Number of sequences")

```

#### Only photosynthetic protists

```{r}

  pr2_genus <- pr2_photo %>% 
  group_by(class, genus) %>% 
  count() %>% 
  ungroup() %>% 
  top_n(30)

  ggplot(pr2_genus) +
    geom_col(aes(x=forcats::fct_reorder(stringr::str_c(class,"-",genus), n), y=n)) +
    coord_flip() +
    ggtitle("Most represented genera - only photosynthetic protists") +
    xlab("Genera") + ylab("Number of sequences")
```


## World sequence distribution

```{r}

map_get_world <- function(resolution="coarse"){
  # Change to "coarse" for global maps / "low" for regional maps
  worldMap <- rworldmap::getMap(resolution = resolution)
  world.points <- fortify(worldMap)
  world.points$region <- world.points$id
  world.df <- world.points[,c("long","lat","group", "region")]
  }


map_world <- function(color_continents = "grey80", 
                      color_borders = "white", 
                      resolution = "coarse") {

  # Background map using the maps package
  # world.df <- map_data("world")

  world.df <- map_get_world(resolution)

  map <- ggplot() +
    geom_polygon(data = world.df, 
                 aes(x=long, y = lat, group = group), 
                 fill=color_continents, 
                 color=color_borders) +
    # scale_fill_manual(values= color_continents , guide = FALSE) +
    scale_x_continuous(breaks = (-4:4) * 45) +
    scale_y_continuous(breaks = (-2:2) * 30) +
    xlab("Longitude") + ylab("Latitude") +
    coord_fixed(1.3) +
    theme_bw()
    # species_map <- species_map + coord_map ()  # Mercator projection
    # species_map <- species_map + coord_map("gilbert") # Nice for the poles
  return(map)
  }
```


### All taxa

```{r map}
 map_world() + geom_point(data=pr2, aes(x=pr2_longitude, y=pr2_latitude), 
                          fill="blue", size=2, shape=21) +
               ggtitle("PR2 - all sequences")  
```

### Photosynthetic protists

```{r}
 map_world() + geom_point(data=pr2_photo, aes(x=pr2_longitude, y=pr2_latitude), 
                          fill="red", size=2, shape=21) +
               ggtitle("PR2 - photosynthetic protists sequences")  
```

