---
title: "Tutorial - Plotting sequence location"
author: "Daniel Vaulot"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{tutorial: plotting sequence location}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  prompt=FALSE,
  tidy=FALSE,
  comment = "#>" 
#  fig.path = "img/"   # Directories where image are saved
)

# Notes
# - cannot use rmarkdown::html_vignette for the moment (maybe with R version 4.0)
# - use build_vignettes() to build the vignettes
# - need to build source package to have the vignettes
```

# Installation

Install from the GitHub web site using the devtools package

```{r, eval=FALSE, message=FALSE}
install.packages("devtools")
devtools::install_github("pr2database/pr2database")
```


# Loading the database


```{r}
library("pr2database")

data("pr2")


```

# Load the libraries


```{r}
  library(dplyr)
  library(ggplot2)    # For plots
```


# Plotting number of sequences per country and ocean

A very good tutorial by [Margaret Mars Brisbin](https://twitter.com/MargaretBrisbin) on how to combine PR2 metadata with Python to locate sequences using all the metadata information (lat, long, country and fuzzy localization): https://maggimars.github.io/eukGeoBlast/eGB.html and https://github.com/maggimars/eukGeoBlast.  This code has been used to incorporate more geo-localisation information into PR2 version 4.12.0



## Number of sequences per country of origin

```{r sequence_country}

  pr2 %>%
    count(pr2_country) %>%
    filter(!is.na(pr2_country) & n > 500) %>%
  ggplot(aes(x = reorder(pr2_country, n), y = n)) +
    geom_col() +
    coord_flip() +
    xlab("") +
    ylab("Number of PR2 sequences")
```

## Number of sequences per ocean of origin

```{r sequence_oceans}

  oceans <- pr2 %>%
    count(pr2_ocean) %>%
    filter(!is.na(pr2_ocean))

  ggplot(oceans, aes(x = reorder(pr2_ocean, n), y = n)) +
    geom_col() +
    coord_flip() +
    xlab("")  + ylab("Number of PR2 sequences")
```
